\mutagen: Coverage-Guided, Property-Based Testing using Exhaustive Type-Preserving Mutations

### 2021-08-23 papers/icse22/main.tex
Pass over last two sections and compression. One page to go.

@@

%% Our solution
In this work, we introduce \mutagen, a CGPT framework that tackles the
aforementioned limitations by applying mutations \emph{exhaustively} (see
Section \ref{sec:mutagen}).
%
%% implemented in Haskell
%% that also builds upon code instrumentation and type-preserving mutations
%
%
% the basic ideas behind \fuzzchick, i.e.,
%
%% \mutagen is introduced in detail in Section \ref{sec:mutagen}.
%
%% Exhaustive mutations
%% Unlike \fuzzchick, mutations in \mutagen are applied \emph{exhaustively}.
%
That is, given an interesting, previously executed test case, our tool
precomputes and schedules every structure-preserving mutation that can be
applied to it.
%
This approach has two inherent advantages. %when compared to \fuzzchick.
%
On one hand, mutations do not rely on randomness to be generated.
%
Instead, every subexpression of the input test case is mutated on the same
basis, which guarantees that no interesting mutation is omitted due to
randomness.
%
%% On the other hand, scheduling mutations exhaustively eliminates the need for a
%% power schedule to assign a given mutation energy to each input test case.
%
On the other hand, scheduling mutations exhaustively eliminates the need for a
power schedule to assign a given mutation budget to each input test case.

@@

%% Heuristics to improve testing loop
Additionally, the testing loop of our tool incorporates two heuristics that help
finding bugs faster and more reliably (Section \ref{sec:heuristics}).
%
In the first place, \mutagen uses first-in first-out (FIFO) scheduling with
priority to enqueue interesting test cases for mutation.
%
This scheduling algorithm is indexed by the novelty of each new test case with
respect to the ones already seen.
%
In this light, interesting test cases that discover new parts of the code
earlier during execution are given a higher priority.
%
Moreover, our scheduling lets the testing loop jump back and forth between
mutable test cases as soon as a new more interesting one is enqueued for
mutation, eliminating potential delays when the mutation queues grow large.

@@

%% Validation
To validate our ideas, we use two main case studies (described in detail in
Section \ref{sec:casestudies}): the Information-Flow Control (IFC) stack machine
used by \citeauthor{lampropoulos2019coverage} in their original work, and an
existing WebAssembly engine of industrial strenght written in Haskell.
%
In both case studies, we additionally compare the effect of the heuristics
implemented on top of the base testing loop of \mutagen.
%
In the IFC stack machine case study, we compare \mutagen against \fuzzchick, the
reference CGPT implementation developed in Coq by
\citeauthor{lampropoulos2019coverage}.
%
Our results (Section \ref{sec:evaluation}) indicate that \mutagen outperforms
\fuzzchick both in terms of time to failure and failure rate.
%
On the other hand, our WebAssembly case study shows the performance of our tool
in a realistic scenario.
%
There, \mutagen is capable of reliably finding 15 planted bugs in the validator
and interpreter, as well as 3 previously unknown bugs that flew under the radar
of the existing unit test suite of this engine.
%
This case study also help us to compare the performance of our tool against a
more traditional PBT approach that does not rely on code instrumentation.
%
All in all, our evaluation encompasses more than 600 hours of computing time and
suggests that testing mutants exhaustively can be an appealing technique for
finding bugs reliably without sacrificing speed.

%% Last two sections
Finally, we discuss related work in Section \ref{sec:related} and conclude
in Section \ref{sec:conclusions}.

@@

The next subsection show how GCPT helps overcoming this limitation without
having to write specialized random generators by hand.

@@

%% Mutators in GCPT use a simple set of mutation operations that are randomly
%% applied to their inputs at the data type level.
%% %
%% In simple terms, these operations encompass
%% %
%% \begin{inparaenum}
%% \item \emph{swapping a data constructor}, replacing its outermost data
%%   constructor by another one, e.g., a \texttt{Leaf} by a \texttt{Branch} and
%%   vice-versa, either dropping unused subexpressions or generating new random
%%   ones when the new data constructor contains extra fields that cannot be filled
%%   using existing ones;
%% %% \item \emph{shrinking the value}, replacing its outermost data constructor with
%% %%   one that contains a subset of its fields, reusing existing subexpressions;
%% %% \item \emph{growing the value}, replacing its outermost data constructor with
%% %%   one that contains a superset of its fields, reusing existing subexpressions
%% %%   and generating random ones when needed;
%% \item \emph{returning a subexpression of the same type}, e.g, returning a left
%%   or right subtree when applied to a \texttt{Branch}; and
%% \item \emph{mutating recursively}, applying a mutation operation over an
%%   immediate subexpression.
%% \end{inparaenum}
%% %
%% We will provide more details about structure-preserving mutations in Section
%% \ref{sec:mutagen}.

@@

All these observations led us to consider three main aspects of CGPT that can be
improved upon --- and that constitute the main goal of this work.
%
%% As mentioned earlier, when we recreated the evaluation of the IFC stack machine
%% case study (described in detail in Section \ref{sec:casestudies}), we found that
%% after 30 runs (as opposed to the 5 runs used original by
%% \citeauthor{lampropoulos2019coverage}), \emph{\fuzzchick was only able to find 5
%%   out of the 20 injected bugs with a failure rate of 1}, the hardest one being
%% found only around 13\% of the time after an hour of testing.
%% %
%% These results are presented in detail in Section \ref{sec:evaluation}.
%
%% At the light of these observations, we identified three main aspects of
%% \fuzzchick that can be improved upon --- and that constitute the main goal of
%% this work.
%
In no particular order:

@@

\item \emph{Enqueuing mutation candidates:}
%
the CGPT testing loop uses two single queues for keeping valid and discarded
mutation candidates.
%
Whenever a new test case is found interesting, it is placed \emph{at the end of
  its corresponding queue}.
%
If this test case happens to have discovered a whole new portion of the target
code, it will not be further mutated until the rest of the queue ahead of it
gets processed.
%
This can limit the effectiveness of the testing loop if the queues tend to grow
more often than they tend to shrink, as interesting mutation candidates can get
``buried'' at the end of a long queue that only exercises the same portion of
the target code.
%
In an extreme case, they might not be processed at all within the testing
budget.
%
Ideally, \emph{one would like a mechanism that prioritizes mutating test cases
  that discover new portions of the code right away}, and that is capable of
jumping back and forth from mutation candidates whenever this happens.
%
%% We show in Section \ref{sec:heuristics} how this can be achieved by analyzing
%% the execution information in order to prioritize test cases with novel execution
%% traces.

@@

In this section we describe the main ideas behind \mutagen, our CGPT tool
written in Haskell.
%
Notably, these ideas extend beyond Haskell and functional programming languages
in general.
%
%% Later, Section \ref{sec:heuristics} introduces heuristics that help finding bugs
%% faster in certain testing scenarios.

In contrast to CGPT, and many other popular fuzzing tools for that matter,
\mutagen works by to mutating test cases on an exhaustive and precise manner,
where
%
\begin{inparaenum}
  \item each subexpression of a mutable test case is associated with a set of
    structure-preserving mutations, and
  \item each one of these mutations is scheduled \emph{exactly once}.
\end{inparaenum}
%
%% There is a small exception to this rules that we will introduce soon.
%
Our realization is that, by following an exhaustive mutation approach, we can
avoid the need for a power schedule, i.e., a heuristic to assign a certain
mutation budget to each mutated test case.
%
Moreover, producing mutants exhaustively ensures that no interesting mutation is
omitted due to randomness, as well a no mutation is evaluated more than once.
%
This is inspired by exhaustive bounded testing tools like \emph{SmallCheck}
\cite{runciman2008smallcheck} or \emph{Korat} \cite{boyapati2002korat}, that
produce test cases exhaustively --- please refer to Section \ref{sec:related}
for a detailed discussion.

@@

%% In simple words, such tools work by enumerating all possible values of the data
%% types used in the testing properties, and then executing them from smaller to
%% larger until a bug is found, or a certain size bound is reached.
%% %
%% The main problem with this approach is that the space of all possible test cases
%% often grows exponentially as we increment the size bound, and the user
%% experiences what it looks like ``hitting a wall'', where no larger test cases
%% can be evaluated until we exhausted all the immediately smaller ones
%% \cite{DuregardJW12}.
%% %
%% To alleviate this problem, such tools often rely on heuristics to prune the
%% search space based on detecting unused subexpressions --- this is discussed in
%% Section \ref{sec:related}.

@@

\begin{figure}[t]
\begin{verbatim}
mutate (Leaf x)       = [ PURE (Branch def x def) ]
mutate (Branch l x r) = [ PURE l, PURE r
                        , PURE (Leaf x)
                        , PURE (Branch l x l)
                        , PURE (Branch r x r)
                        , PURE (Branch r x l) ]
\end{verbatim}
%% \vspace{-5pt}
\caption{\label{fig:mutagen:mutator}\mutagen mutator for the \texttt{Tree} data
  type.}
%% \vspace{-10pt}
\end{figure}

@@

\begin{equation}
  \nonumber
  {\small \texttt{positions}}\quad
  \left(
  \tikz[
    baseline=-7mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{Branch}}
      child {node {\small \texttt{Leaf}}
        child {node {\small \texttt{1}}}
      }
      child {node {\small \texttt{2}}}
      child {node {\small \texttt{Leaf}}
        child {node {\small \texttt{3}}}
      };
  }
  \right)
  \quad
  =
  \quad
  \tikz[
    baseline=-7mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\texttt{\small []}}
      child {node {\small \texttt{[0]}}
        child {node {\small \texttt{[0,0]}}}
      }
      child {node {\small \texttt{[1]}}}
      child {node {\small \texttt{[2]}}
        child {node {\small \texttt{[2,0]}}}
      };
  }
\end{equation}

@@

\begin{figure*}[t]
%% \vspace{-15pt}
\newcommand\stacked[2]{\underset{\textstyle #2\mathstrut}{#1}}
\begin{align*}
  \nonumber
  \qquad\qquad\qquad\qquad
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
        }
      }
  }
  \right]
  \;
  +
  \;
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{5}}}
        }
      }
  }
  \;
  =
  \;
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\color{red} \small \texttt{1}}
      child {node {\color{red} \small \texttt{2}}
        child {node {\color{red} \small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\color{blue} \small \texttt{5}}}
        }
      }
  }
  \right]
  ,
  \stacked{{\color{blue} new=1\phantom{x}}}{{\color{red} depth=3}}
  %% \end{equation}
  %% \begin{equation}
  %% \nonumber
  %% \\
  \qquad
  \qquad
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\small \texttt{5}}}
        }
      }
  }
  \right]
  \;
  +
  \;
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{6}}
          child {node {\small \texttt{7}}
          }
        }
      }
  }
  \;
  =
  \;
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=1cm}
  ]{
    \node {\color{red} \small \texttt{1}}
      child {node {\color{red} \small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\small \texttt{5}}}
        }
        child {node {\color{blue}\small \texttt{6}}
          child {node {\color{blue}\small \texttt{7}}}
        }
      }
  }
  \right]
  ,
  \stacked{{\color{blue} new=2\phantom{x}}}{{\color{red} depth=2}}
\end{align*}
%% \vspace{-10pt}
\caption{\label{fig:mutagen:tracelog}Inserting two new execution traces into the
  internal trace log (represented using brackets).}
\end{figure*}

@@

\begin{algorithm}[b]
  \SetArgSty{textnormal}
  %% \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, R, gen, mut}}{
    $\cdots$\;
    x $\gets$ Pick(QValid, QDisc, gen, mut)\;
    (result, trace) $\gets$ WithTrace(P(x))\;
    $\cdots$\;
    \If{Interesting(TLog, trace)}{
      \If{\KwNot Discarded(result)}{
        batch $\gets$ Mutate(x, mut, R)\;
        {\color{red} prio $\gets$ BranchDepth(TLog, trace)}\; %% NEW
        {\color{red} PushFront(QValid, prio, batch)}\;          %% NEW
      }
      $\cdots$\;
    }
  }
  %% \vspace{5pt}
  \SetKwFunction{Pick}{Pick}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\Pick{QValid, QDisc, gen}}{
    \If{\KwNot Empty(QValid)}{
      {\color{red} (batch, prio) $\gets$ DequeMin(QValid)}\; %%NEW
      \If{Empty(batch)}{
        Pick(QValid, QDisc, gen)\;
      } \uElse{
        {\color{red} PushFront(QValid, prio, Rest(batch))}\; %% NEW
        \KwRet First(batch)\;
      }
    }
    $\cdots$\;
  }

@@

@@

As introduced in Section \ref{sec:mutagen}, our tool is parameterized by the
number of random mutations to be generated over each mutable subexpression
defined using a random mutant, e.g., numeric values.
%
But, how many random mutations should we use? A single one? A few tenths? A few
hundred?
%% %
%% Using too little can put finding bugs at risk.
%% %
%% For instance, when the system under test branches based on numeric values, we
%% should make sure that we set enough random mutations to test each branch on a
%% reasonable basis.
%% %
%% Using too many, on the other hand, can degrade the performance of the testing
%% loop, as it will spend too much time producing uninteresting mutations.
%% %
%% This can happen for instance if the subexpressions defined using random mutants
%% are only used as payloads, and their value does not affect the execution in any
%% way.
%
Answering this question precisely is not an easy task, and this second heuristic
aims to tackle this issue.

@@

\begin{center}
\footnotesize
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Instruction}
& \textbf{Precondition Check}
& \textbf{Final PC Label}
& \textbf{Final Result Label} \\
\hline
\texttt{Store}
& $l_{pc} \vee l_{p} \sqsubseteq l_{v}$
& $l_{pc}$
& $l_{v'} \vee l_{pc} \vee l_{p}$ \\
\hline
\end{tabular}
\end{center}

@@

\begin{table}[t]
\footnotesize
\begin{tabular}{|c|c|l|}
\hline
\textbf{Id}
& \textbf{Subsystem}
& \textbf{Description} \\
\hline
1
& Validator
& Wrong if-then-else type validation on else branch \\
\hline
2
& Validator
& Wrong stack type validation \\
\hline
3
& Validator
& Removed function type mismatch assertion \\
\hline
4
& Validator
& Removed max memory instances assertion \\
\hline
5
& Validator
& Removed function index out-of-range assertion \\
\hline
6
& Validator
& Wrong type validation on \texttt{i64.eqz} instruction \\
\hline
7
& Validator
& Wrong type validation on \texttt{i32} binary operations \\
\hline
8
& Validator
& Removed memory index out-of-range assertion \\
\hline
9
& Validator
& Wrong type validation on \texttt{i64} constants \\
\hline
10
& Validator
& Removed alignment validation on \texttt{i32.load} instruction \\
\hline
11
& Interpreter
& Wrong interpretation of \texttt{i32.sub} instruction \\
\hline
12
& Interpreter
& Wrong interpretation of \texttt{i32.lt\_u} instruction \\
\hline
13
& Interpreter
& Wrong interpretation of \texttt{i32.shr\_u} instruction \\
\hline
14
& Interpreter
& Wrong local variable initialization \\
\hline
15
& Interpreter
& Wrong memory address casting on \texttt{i32.load8\_s} instruction \\
\hline
\end{tabular}
\caption{\label{table:wasm:injectedbugs}Bugs injected into
  \textit{haskell-wasm}.}
%% \vspace{-15pt}
\end{table}

@@

In both case studies, we show how the FIFO scheduling and trace reset heuristics
described in Section \ref{sec:heuristics} affect the testing performance of
\mutagen by individually disabling them.
%
We call these variants \textit{no FIFO} and \textit{no reset}, respectively.
%
In the case of \textit{no reset}, the amount of random mutations is no longer
controlled by this heuristic, so we arbitrarily fixed it to 25 random mutations
throughout the execution of the tool.

>>>>

\setlength{\textfloatsep}{5pt}
\setlength{\floatsep}{5pt}

@@

%% Our solution
In this work, we introduce \mutagen, a CGPT framework that tackles the
aforementioned limitations by applying mutations \emph{exhaustively} (see
\S \ref{sec:mutagen}).
%
%% implemented in Haskell
%% that also builds upon code instrumentation and type-preserving mutations
%
%
% the basic ideas behind \fuzzchick, i.e.,
%
%% \mutagen is introduced in detail in \S \ref{sec:mutagen}.
%
%% Exhaustive mutations
%% Unlike \fuzzchick, mutations in \mutagen are applied \emph{exhaustively}.
%
That is, given an interesting, previously executed test case, our tool
precomputes and schedules every structure-preserving mutation that can be
applied to it.
%
This approach has two inherent advantages. %when compared to \fuzzchick.
%
On one hand, mutations do not rely on randomness to be generated.
%
Instead, every subexpression of the input test case is mutated on the same
basis, which guarantees that no interesting mutation is omitted due to
randomness.
%
%% On the other hand, scheduling mutations exhaustively eliminates the need for a
%% power schedule to assign a given mutation energy to each input test case.
%
On the other hand, scheduling mutations exhaustively eliminates the need for a
power schedule to assign a given mutation budget to each input test case.

@@

%% Heuristics to improve testing loop
Additionally, the testing loop of our tool incorporates two heuristics that help
finding bugs faster and more reliably (\S \ref{sec:heuristics}).
%
In the first place, \mutagen uses first-in first-out (FIFO) scheduling with
priority to enqueue interesting test cases for mutation.
%
This scheduling algorithm is indexed by the novelty of each new test case with
respect to the ones already seen.
%
In this light, interesting test cases that discover new parts of the code
earlier during execution are given a higher priority.
%
Moreover, our scheduling lets the testing loop jump back and forth between
mutable test cases as soon as a new more interesting one is enqueued for
mutation, eliminating potential delays when the mutation queues grow large.

@@

%% Validation
To validate our ideas, we use two main case studies (described in detail in
\S \ref{sec:casestudies}): the Information-Flow Control (IFC) stack machine
used by \citeauthor{lampropoulos2019coverage} in their original work, and an
existing WebAssembly engine of industrial strenght written in Haskell.
%
In both case studies, we additionally compare the effect of the heuristics
implemented on top of the base testing loop of \mutagen.
%
In the IFC stack machine case study, we compare \mutagen against \fuzzchick, the
reference CGPT implementation developed in Coq by
\citeauthor{lampropoulos2019coverage}.
%
Our results (\S \ref{sec:evaluation}) indicate that \mutagen outperforms
\fuzzchick both in terms of time to failure and failure rate.
%
On the other hand, our WebAssembly case study shows the performance of our tool
in a realistic scenario.
%
There, \mutagen is capable of reliably finding 15 planted bugs in the validator
and interpreter, as well as 3 previously unknown bugs that flew under the radar
of the existing unit test suite of this engine.
%
This case study also help us to compare the performance of our tool against a
more traditional PBT approach that does not rely on code instrumentation.
%
All in all, our evaluation encompasses more than 600 hours of computing time and
suggests that testing mutants exhaustively can be an appealing technique for
finding bugs reliably without sacrificing speed.

%% Last two sections
Finally, we discuss related work in \S \ref{sec:related} and conclude
in \S \ref{sec:conclusions}.

@@

The next subsection shows how GCPT helps overcoming this limitation without
having to write specialized generators by hand.

@@

%% Mutators in GCPT use a simple set of mutation operations that are randomly
%% applied to their inputs at the data type level.
%% %
%% In simple terms, these operations encompass
%% %
%% \begin{inparaenum}
%% \item \emph{swapping a data constructor}, replacing its outermost data
%%   constructor by another one, e.g., a \texttt{Leaf} by a \texttt{Branch} and
%%   vice-versa, either dropping unused subexpressions or generating new random
%%   ones when the new data constructor contains extra fields that cannot be filled
%%   using existing ones;
%% %% \item \emph{shrinking the value}, replacing its outermost data constructor with
%% %%   one that contains a subset of its fields, reusing existing subexpressions;
%% %% \item \emph{growing the value}, replacing its outermost data constructor with
%% %%   one that contains a superset of its fields, reusing existing subexpressions
%% %%   and generating random ones when needed;
%% \item \emph{returning a subexpression of the same type}, e.g, returning a left
%%   or right subtree when applied to a \texttt{Branch}; and
%% \item \emph{mutating recursively}, applying a mutation operation over an
%%   immediate subexpression.
%% \end{inparaenum}
%% %
%% We will provide more details about structure-preserving mutations in \S
%% \ref{sec:mutagen}.

@@

All these observations led us to consider three main aspects of CGPT that can be
improved upon --- and that constitute the main goal of this work.
%
%% As mentioned earlier, when we recreated the evaluation of the IFC stack machine
%% case study (described in detail in \S \ref{sec:casestudies}), we found that
%% after 30 runs (as opposed to the 5 runs used original by
%% \citeauthor{lampropoulos2019coverage}), \emph{\fuzzchick was only able to find 5
%%   out of the 20 injected bugs with a failure rate of 1}, the hardest one being
%% found only around 13\% of the time after an hour of testing.
%% %
%% These results are presented in detail in \S \ref{sec:evaluation}.
%
%% At the light of these observations, we identified three main aspects of
%% \fuzzchick that can be improved upon --- and that constitute the main goal of
%% this work.
%
In no particular order:

@@

\item \emph{Enqueuing mutation candidates:}
%
the CGPT testing loop uses two single queues for keeping valid and discarded
mutation candidates.
%
Whenever a new test case is found interesting, it is placed \emph{at the end of
  its corresponding queue}.
%
If this test case happens to have discovered a whole new portion of the target
code, it will not be further mutated until the rest of the queue ahead of it
gets processed.
%
This can limit the effectiveness of the testing loop if the queues tend to grow
more often than they tend to shrink, as interesting mutation candidates can get
``buried'' at the end of a long queue that only exercises the same portion of
the target code.
%
In an extreme case, they might not be processed at all within the testing
budget.
%
Ideally, \emph{one would like a mechanism that prioritizes mutating test cases
  that discover new portions of the code right away}, and that is capable of
jumping back and forth from mutation candidates whenever this happens.
%
%% We show in \S \ref{sec:heuristics} how this can be achieved by analyzing
%% the execution information in order to prioritize test cases with novel execution
%% traces.

@@

In this section we describe the main ideas behind \mutagen, our CGPT tool
written in Haskell.
%
Notably, these ideas extend beyond Haskell and functional programming languages
in general.
%
%% Later, \S \ref{sec:heuristics} introduces heuristics that help finding bugs
%% faster in certain testing scenarios.

In contrast to CGPT, and many other popular fuzzing tools for that matter,
\mutagen works by to mutating test cases on an exhaustive and precise manner,
where
%
\begin{inparaenum}
  \item each subexpression of a mutable test case is associated with a set of
    structure-preserving mutations, and
  \item each one of these mutations is scheduled \emph{exactly once}.
\end{inparaenum}
%
%% There is a small exception to this rules that we will introduce soon.
%
Our realization is that, by following an exhaustive mutation approach, we can
avoid the need for a power schedule, i.e., a heuristic to assign a certain
mutation budget to each mutated test case.
%
Moreover, producing mutants exhaustively ensures that no interesting mutation is
omitted due to randomness, as well a no mutation is evaluated more than once.
%
This is inspired by exhaustive bounded testing tools like \emph{SmallCheck}
\cite{runciman2008smallcheck} or \emph{Korat} \cite{boyapati2002korat}, that
produce test cases exhaustively --- please refer to \S \ref{sec:related}
for a detailed discussion.

@@

%% In simple words, such tools work by enumerating all possible values of the data
%% types used in the testing properties, and then executing them from smaller to
%% larger until a bug is found, or a certain size bound is reached.
%% %
%% The main problem with this approach is that the space of all possible test cases
%% often grows exponentially as we increment the size bound, and the user
%% experiences what it looks like ``hitting a wall'', where no larger test cases
%% can be evaluated until we exhausted all the immediately smaller ones
%% \cite{DuregardJW12}.
%% %
%% To alleviate this problem, such tools often rely on heuristics to prune the
%% search space based on detecting unused subexpressions --- this is discussed in
%% \S \ref{sec:related}.

@@

\begin{figure}[t]
\begin{verbatim}
mutate (Leaf x)       = [ PURE (Branch def x def) ]
mutate (Branch l x r) = [ PURE l, PURE r
                        , PURE (Leaf x)
                        , PURE (Branch l x l)
                        , PURE (Branch r x r)
                        , PURE (Branch r x l) ]
\end{verbatim}
\vspace{-5pt}
\caption{\label{fig:mutagen:mutator}\mutagen mutator for the \texttt{Tree} data
  type.}
%% \vspace{-15pt}
\end{figure}

@@

\begin{equation}
  \nonumber
  \footnotesize
  \texttt{positions}\quad
  \left(
  \tikz[
    baseline=-6mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.5cm}
  ]{
    \node {\texttt{Branch}}
      child {node {\texttt{Leaf}}
        child {node {\texttt{1}}}
      }
      child {node {\texttt{2}}}
      child {node {\texttt{Leaf}}
        child {node {\texttt{3}}}
      };
  }
  \right)
  \quad
  =
  \quad
  \tikz[
    baseline=-6mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.5cm}
  ]{
    \node {\texttt{[]}}
      child {node {\texttt{[0]}}
        child {node {\texttt{[0,0]}}}
      }
      child {node {\texttt{[1]}}}
      child {node {\texttt{[2]}}
        child {node {\texttt{[2,0]}}}
      };
  }
\end{equation}

@@

@@

\begin{algorithm}[t]
  \SetArgSty{textnormal}
  %% \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, R, gen, mut}}{
    $\cdots$\;
    x $\gets$ Pick(QValid, QDisc, gen, mut)\;
    (result, trace) $\gets$ WithTrace(P(x))\;
    $\cdots$\;
    \If{Interesting(TLog, trace)}{
      \If{\KwNot Discarded(result)}{
        batch $\gets$ Mutate(x, mut, R)\;
        {\color{red} prio $\gets$ BranchDepth(TLog, trace)}\; %% NEW
        {\color{red} PushFront(QValid, prio, batch)}\;          %% NEW
      }
      $\cdots$\;
    }
  }
  %% \vspace{5pt}
  \SetKwFunction{Pick}{Pick}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\Pick{QValid, QDisc, gen}}{
    \If{\KwNot Empty(QValid)}{
      {\color{red} (batch, prio) $\gets$ DequeMin(QValid)}\; %%NEW
      \If{Empty(batch)}{
        Pick(QValid, QDisc, gen)\;
      } \uElse{
        {\color{red} PushFront(QValid, prio, Rest(batch))}\; %% NEW
        \KwRet First(batch)\;
      }
    }
    $\cdots$\;
  }

@@

\begin{figure}[t]
%% \vspace{-15pt}
\scriptsize
\newcommand\stacked[2]{\underset{\textstyle #2\mathstrut}{#1}}
\begin{align*}
  \nonumber
  %% \qquad\qquad\qquad\qquad
  \left[
  \tikz[
    baseline=-8mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.5cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
        }
      }
  }
  \right]
  \;
  +
  \;
  \tikz[
    baseline=-8mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.5cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{5}}}
        }
      }
  }
  \;
  =
  \;
  \left[
  \tikz[
    baseline=-8mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.5cm}
  ]{
    \node {\color{red} \small \texttt{1}}
      child {node {\color{red} \small \texttt{2}}
        child {node {\color{red} \small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\color{blue} \small \texttt{5}}}
        }
      }
  }
  \right]
  ,
  \stacked{{\color{blue} new=1\phantom{x}}}{{\color{red} depth=3}}
  %% \end{equation}
  %% \begin{equation}
  %% \nonumber
  %% \\
  %% \qquad
  \quad
  \left[
  \tikz[
    baseline=-8mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.5cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\small \texttt{5}}}
        }
      }
  }
  \right]
  \;
  +
  \;
  \tikz[
    baseline=-8mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.5cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{6}}
          child {node {\small \texttt{7}}
          }
        }
      }
  }
  \;
  =
  \;
  \left[
  \tikz[
    baseline=-8mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\color{red} \small \texttt{1}}
      child {node {\color{red} \small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\small \texttt{5}}}
        }
        child {node {\color{blue}\small \texttt{6}}
          child {node {\color{blue}\small \texttt{7}}}
        }
      }
  }
  \right]
  ,
  \stacked{{\color{blue} new=2\phantom{x}}}{{\color{red} depth=2}}
\end{align*}
\vspace{-10pt}
\caption{\label{fig:mutagen:tracelog}Inserting two new execution traces into the
  internal trace log (represented using brackets).}
\end{figure}

@@

As introduced in \S \ref{sec:mutagen}, our tool is parameterized by the
number of random mutations to be generated over each mutable subexpression
defined using a random mutant, e.g., numeric values.
%
But, how many random mutations should we use? A single one? A few tenths? A few
hundred?
%% %
%% Using too little can put finding bugs at risk.
%% %
%% For instance, when the system under test branches based on numeric values, we
%% should make sure that we set enough random mutations to test each branch on a
%% reasonable basis.
%% %
%% Using too many, on the other hand, can degrade the performance of the testing
%% loop, as it will spend too much time producing uninteresting mutations.
%% %
%% This can happen for instance if the subexpressions defined using random mutants
%% are only used as payloads, and their value does not affect the execution in any
%% way.
%
Answering this question precisely is not an easy task, and this second heuristic
aims to tackle this issue.

@@

\begin{center}
\scriptsize
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Instruction}
& \textbf{Precondition Check}
& \textbf{Final PC Label}
& \textbf{Final Result Label} \\
\hline
\texttt{Store}
& $l_{pc} \vee l_{p} \sqsubseteq l_{v}$
& $l_{pc}$
& $l_{v'} \vee l_{pc} \vee l_{p}$ \\
\hline
\end{tabular}
\end{center}

@@

\begin{table}[t]
\scriptsize
\begin{tabular}{|c|c|l|}
\hline
\textbf{Id}
& \textbf{Subsystem}
& \textbf{Description} \\
\hline
1
& Validator
& Wrong if-then-else type validation on else branch \\
\hline
2
& Validator
& Wrong stack type validation \\
\hline
3
& Validator
& Removed function type mismatch assertion \\
\hline
4
& Validator
& Removed max memory instances assertion \\
\hline
5
& Validator
& Removed function index out-of-range assertion \\
\hline
6
& Validator
& Wrong type validation on \texttt{i64.eqz} instruction \\
\hline
7
& Validator
& Wrong type validation on \texttt{i32} binary operations \\
\hline
8
& Validator
& Removed memory index out-of-range assertion \\
\hline
9
& Validator
& Wrong type validation on \texttt{i64} constants \\
\hline
10
& Validator
& Removed alignment validation on \texttt{i32.load} instruction \\
\hline
11
& Interpreter
& Wrong interpretation of \texttt{i32.sub} instruction \\
\hline
12
& Interpreter
& Wrong interpretation of \texttt{i32.lt\_u} instruction \\
\hline
13
& Interpreter
& Wrong interpretation of \texttt{i32.shr\_u} instruction \\
\hline
14
& Interpreter
& Wrong local variable initialization \\
\hline
15
& Interpreter
& Wrong memory address casting on \texttt{i32.load8\_s} instruction \\
\hline
\end{tabular}
\caption{\label{table:wasm:injectedbugs}Bugs injected into
  \textit{haskell-wasm}.}
%% \vspace{-15pt}
\end{table}

@@

In both case studies, we show how the FIFO scheduling and trace reset heuristics
described in \S \ref{sec:heuristics} affect the testing performance of
\mutagen by individually disabling them.
%
We call these variants \textit{no FIFO} and \textit{no reset}, respectively.
%
In the case of \textit{no reset}, the amount of random mutations is no longer
controlled by this heuristic, so we arbitrarily fixed it to 25 random mutations
throughout the execution of the tool.
